---
title: "Basic Data Loading and Transformation"
author: "Angela Zoss"
date: "`r Sys.Date()`"
output: html_document
---

```{r}

# Run the first time you start working with this file

install.packages("tidyverse")


```

```{r}

# Loading required packages

library(tidyverse)

```

## Loading data

The `tidyverse` comes with some built-in data loading functions that I recommend
over the ones that are included in base R. The ones I use most commonly are `read_csv`
and `read_xlsx`, which is actually in a separate package called `readxl`. 

Both of these functions can load data either from a local data file or from something
shared on the web. Let's grab the dataset from the Open Refine lesson for Library Carpentries.

```{r}

lc_open_refine <- read_csv(
  'https://github.com/LibraryCarpentry/lc-open-refine/raw/gh-pages/data/doaj-article-sample.csv',
  col_types = cols(.default = "c", Date = col_date(format="%m/%d/%Y")) 
  )

```

The `read_csv()` function does a few things automatically. It assumes that the first 
row of the dataset contained the column names. It can also each column to try to guess 
the data type. Usually the tidyverse data reading functions do a great job of guessing
data types, but you can also specify them manually if you need to. Here, I specified
that Date is a date but all others are character variables.

## Explore a dataset

To see the contents of a variable, we can start by just typing the name of the variable.

```{r, echo=TRUE}

lc_open_refine

```

With data frames, though, there are a couple of additional functions that can be helpful.
Try the following commands to see what kind of information they provide.

### Summarize a dataset

```{r, echo=TRUE}

summary(lc_open_refine)

```

### Glimpse a dataset

```{r, echo=TRUE}

glimpse(lc_open_refine)

```

Now, let's say we've identified a few columns that need a bit of data processing.
Let's walk through working on a few specific columns.

## `select()` for isolating columns

If you've already learned about accessing data using base R capabilities, you may
have used `$` or `[]` or `[[]]` to grab one of the columns in a data frame. Instead of 
teaching those, I'm going to skip ahead to the tidyverse functions.

The `select()` function has a wide variety of ways to specify the columns you want.
Let's start with names. To select by name, all you have to do is list all of the
column names you want, separated by commas. The first argument in the function
will always be the name of the data frame.

Try selecting the DOI column.

```{r}

select(lc_open_refine, DOI)

```

Now, let's pretend we need to remove any rows where the DOI is empty.

## Find and remove rows with missing DOIs

In `tidyverse`, the function for retrieving a set of rows is `filter()`. Usually, 
if you're limiting a dataset to a subset of rows, you want to do it on some specific
criteria, usually related to the data.


### Step 1: Isolating groups of rows by a condition

The general pattern is:

`filter(data_frame, some test for the rows)`

How to test if values are null?

`is.na(name_of_column)`

Combine these to find the NA values in the DOI column:

```{r, echo=TRUE}

filter(lc_open_refine, is.na(DOI))

```

But we want to remove the NAs, not all of the other rows. How do we get the opposite?

We can use `!` to get the negative of a test.

`!is.na(name_of_column)`

Or, in this case:

```{r, echo=TRUE}

filter(lc_open_refine, !is.na(DOI))

```

The tidyverse actually has a handy function that does this: drop_na()

```{r, echo=TRUE}

drop_na(lc_open_refine, DOI)

```

## Fix inconsistencies in Language

When we're exploring a new data set, we might want to ask questions like: What are the unique values of a column? How many records are there for each value?

Tidyverse also has a handy function for this: `count()`.

`count(data_frame, name_of_column)`

Let's use `count()` to summarize the Language column.

```{r, echo=TRUE}

count(lc_open_refine, Language)

```

This shows an inconsistency in some of the records. We need to edit the data
values to standardize them.

### How to edit a variable: mutate

The `mutate()` function allows you to make changes to an existing variable or to
create a new variable. The general pattern is:

`mutate(data_frame, name_of_column = new value or expression)`

We know we need to mutate the Language column, but we have to come up with an 
expression that will do the right mutation. What should the expression do?

* If the Language is "English", change to "EN"
* Otherwise, leave the original value

There are multiple ways we could accomplish this in R. Here are a couple of options:

* Use a simple if/then statement to look for "English" and replace it when we find
* Use a simple string replace to look for "English" and replace it when we find

### Option 1: If/then

R has a more complicated way of writing if-then statements. The `ifelse()` function 
is a simple version that works really well with the mutate function. The general 
pattern is:

`ifelse(test, what to do if true, what to do if false)`

So, in this case, we will want a statement like this:

`ifelse(Language == "English", "EN", Language)`

### Option 2: String replacement

String replacement is similar, but it's more flexible. The function `str_replace()`
will look for English, even if it's not the full contents of the cell in the table.

`str_replace(text, what to replace, what to use instead)`

In our case, we'll want a statement like this:

`str_replace(Language, "English", "EN")`

Now that we have our expression, we need to plug it into the `mutate()` function.

```{r, echo=TRUE}

mutate(lc_open_refine, Language = ifelse(Language == "English", "EN", Language))

```

Or, if we're using option 2:

```{r, echo=TRUE}

mutate(lc_open_refine, Language = str_replace(Language, "English", "EN"))

```

## Fix inconsistencies in Publisher

Next let's check out the Publisher column. Try using `count()` to explore the categories.

```{r, echo=TRUE}

count(______)

```

What needs to be fixed? Can you think of a string replacement that will work?


```{r, echo=TRUE}

mutate(lc_open_refine, Publisher = str_replace(Publisher, "", ""))

```

## Chaining operations

### The magrittr pipe

`%>%`

This symbol is called a pipe, and it was originally developed for a package
called magrittr. (Many R packages end with a single r, rather than "er.")

The pipe works by taking whatever the result of the previous command is and injecting
or piping it straight into the next command. Let's start with a simple example.

```{r, echo=TRUE}

class("a")

"a" %>% class()

```

class() is a basic function that returns the class of a data element in R: `class()`.
Inside the parentheses, you would normally type the variable name or the data element
you want to find the class of. If you want to use the pipe, though, you could start
with the data element and then pipe it to the `class()` function.

Those two statements are identical. When you use the pipe, whatever was on the left
side of the pipe gets inserted into the parentheses of the function on the right.

## Create a single chain of data transformations to remove null DOIs and fix the Language and Publisher columns

First, let's take the data set and pipe it to a function that will drop the NAs in 
the DOI column.

```{r, echo=TRUE}

lc_open_refine %>% drop_na(DOI)

```

Next, add another step to change "English" to "EN" in Language.

```{r, echo=TRUE}

lc_open_refine %>% drop_na(DOI) %>% mutate(Language = str_replace(Language, "English", "EN"))

```

To avoid long lines, you can start a new line after a pipe.

```{r, echo=TRUE}

lc_open_refine %>% 
  drop_na(DOI) %>% 
  mutate(Language = str_replace(Language, "English", "EN"))

```

Finally, add a link that will fix the Publisher space problem.

```{r, echo=TRUE}

lc_open_refine %>% 
  drop_na(DOI) %>% 
  mutate(Language = str_replace(Language, "English", "EN")) %>%
  ___________

```

The pipeline is working, but the transformed data aren't getting saved anywhere,
they're just getting printed. Let's pick a new name and save the whole pipeline
to that new variable.

```{r, echo=TRUE}

lc_open_refine_transformed <- lc_open_refine %>% 
  ____________ %>%
  ____________ %>%
  ____________

```

## Export the transformed data

After a successful data transformation sequence, you may want to save the file out.
In the tidyverse, we can do this easily with `write_csv()`.


```{r, echo=TRUE, eval=FALSE}

write_csv(lc_open_refine_transformed, "lc_open_refine_transformed.csv")

```

A note of warning, though: NAs are saved in the file as "NA" by default. I often
change these to empty string "" for simplicity.

```{r, echo=TRUE, eval=FALSE}

write_csv(lc_open_refine_transformed, "lc_open_refine_transformed.csv", na="")

```


