---
title: "Working with Multiple Tables"
author: "Angela Zoss"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
---

```{r}

# Run the first time you start working with this file

install.packages("tidyverse")


```

```{r}

# Loading required packages

library(tidyverse)

```

## Loading data

Let's load the dataset from the SQL lesson for Library Carpentries. (I downloaded
the original data, loaded it into the SQLite software, then exported each table
as a separate .csv file.)

```{r, message=FALSE}

articles <- read_csv("data_files/articles.csv")
journals <- read_csv("data_files/journals.csv")
languages <- read_csv("data_files/languages.csv")
licences <- read_csv("data_files/licences.csv")
publishers <- read_csv("data_files/publishers.csv")

```

## Select column(s)

Review the `select()` function from Exercise 1. Try selecting a few columns
from the articles table, like `Title` and `Citation_Count`.

```{r, eval=F}

articles %>% 
  select(_____)

```

## Sorting the results

In the tidyverse, you can sort a table by using the function `arrange()` and then
specifying what column you want to sort by. The `Citation_Count` column might be
a good place to start. If we limit to just a few columns, it will be easier to
see the results.

```{r, eval=F}

articles %>% 
  select(Title, Citation_Count) %>%
  arrange(_____)

```

What if we want the sort to start with the largest values? In SQL, we used this syntax:

```
SELECT *
FROM table name
ORDER BY column_name_1 DESC;
```

In tidyverse, we can surround the name of a column with the helper function `desc()`
to reverse the sort on that column.

```{r}

articles %>% 
  select(Title, Citation_Count) %>%
  arrange(desc(Citation_Count))

```

## Challenge #1

Write a query that returns `Title`, `First_Author`, `ISSNs` and `Citation_Count` from the articles table, ordered by the top cited article and alphabetically by title.

```{r}

```

## Filtering

In SQL, we used the `WHERE` keyword to limit our table to a specific
set of rows. In tidyverse, we use the `filter()` command for that. The
ways we specify our tests, though, will be pretty similar. 

This general SQL example:

```
SELECT *
FROM table name
WHERE column_name_1 = 'some value';
```

would look like this in R:

```
table %>%
  filter(column_name_1 == 'some value')
```

Try to filter articles to just the rows where `Citation_Count` is exactly 10.

```{r, eval=F}

articles %>% 
  select(Title, Citation_Count) %>%
  filter(_____)

```

In SQL, combining multiple tests looks like this:

```
SELECT *
FROM table name
WHERE (column_name_1 = 'some value') AND
(column_name_2 = 'some other value');
```

In tidyverse, you just use `&` to combine tests inside the
`filter()`.

```
table %>%
  filter(column_name_1 == 'some value' & 
         column_name_2 == 'some other value')
```

The equivalent to SQL's `OR` in tidyverse is a single `|` character.

```
table %>%
  filter(column_name_1 == 'some value' | 
         column_name_2 == 'some other value')
```

### Fuzzy Filters

In SQL, we learned about using `LIKE` to add a wildcard into our filter. When we
used `LIKE`, we could add `%` into the search term to indicate the wildcard.

```
SELECT *
FROM table name
WHERE column_name_1 LIKE ‘%part of value%’;
```

In tidyverse, there are various string functions that can be used to search for
substrings. One similar approach would be used a function called `str_detect()`. 
This function works like `str_replace()`, which we learned in the previous exercise.
The first argument is the column name, and then you include the string sequence 
you're looking for.

For example, let's look for titles that contain "Classification."

```{r}

articles %>%
  select(Title, Citation_Count) %>%
  filter(str_detect(Title, "Classification"))

```

I highly recommend browsing through all of the `str_` functions in the `stringr` package. 
You can go to the Packages tab and then click on `stringr` to see the package 
functions in a list.

## Challenge #2

Write a query that returns `Title`, `First_Author`, `Subjects`, `ISSNs`, `Month` and `Year` for all papers where `Subjects` contains “computer” and that have more than 8 citations.

## Aggregation

Aggregation and calculation go together. When you want to summarize across
multiple rows in a table, you first have to aggregate those rows into a
group, and then you have specify what calculation you want to apply to the
values in that column.

In SQL, you specify the columns you want to group by, and then you apply
aggregation functions to the remaining columns. 

```
SELECT column_name_1, AVG(column_name_2)
FROM table name
GROUP BY column_name_1;
```

The same is true in tidyverse. There is a `group_by()` function that takes
a list of columns. After you group, you then use `summarise()` to specify
aggregations for any ungrouped columns you want to keep. You don't have to explicitly
select the grouped columns for them to show up in the result.

The `summarise()` function works like `mutate()`. The first thing you write is
the name you want the new column to have, then you type `=`, then you specify
how you want to aggregate one of the original columns.

```{r}

articles %>%
  group_by(First_Author) %>%
  summarise(total_citations = sum(Citation_Count))


```

## Challenge #3

Write a query using an aggregate function that returns the number of article titles per ISSNs, sorted by title count in descending order.

(Hint: in SQL, we had a function named `COUNT`. In tidyverse, the equivalent function is `n()`.)

```{r}



```

Note: aggregation functions can get more sophisticated than just a count or an average.
Check out the dplyr cheat sheet (Help -> Cheat Sheets -> Data Transformation with dplyr)
for more options.

## Challenge #4

Write a query that returns, from the articles table, the average `Citation_Count` for each journal ISSN, but only for the journals with 5 or more citations on average.

```{r}

```

## Joins

Databases really shine when it comes to blending data from multiple
tables. Being able to do this in R is where you can really move past
what's easy to do in, say, OpenRefine.

In SQL, a basic join is:

```
SELECT *
FROM table_name_1
JOIN table_name_2
ON table_name_1.column_name = table_name_2.column_name;
```

In tidyverse, you start with one table and then use the 
`left_join()` function to bring in the second table. 

If there are matching column names, they'll be matched automatically.
If you want to specify which columns should be matched, you can add 
`by = join_by(column_in_first_table == column_in_second_table)`

```{r}

articles %>% 
  left_join(languages, by = join_by(LanguageId == id))

```

## Challenge #5

Write a query that JOINS the articles and journals tables and that returns the Journal_Title, total number of articles published and average number of citations for every journal ISSN.

```{r}

```

## Challenge #6

Write a query that returns the Journal_Title, Publisher name, and number of articles published, ordered by number of articles in descending order.

```{r}

```

