---
title: "R Data Types"
author: "Angela Zoss"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r}

# Loading required packages

library(tidyverse)

```

## How to write R code

R is a programming language that was originally designed for statistics.
It's an old enough language that there are a bunch of variations to how
you can accomplish any particular task.

A couple of things are pretty consistent, though. Within any block of
code you are running, R code is executed line by line, top to bottom.
Also, R will either expect the full command to be on the same line, or
you will have to be careful about where to add a line break. You are
usually safe adding a line break after a comma, a left parenthesis `(`,
or a left curly brace `{`.

This won't matter too much for our simple examples, but when you have
longer and longer operations, you will definitely want to break them up
over multiple lines for readability.

The simplest R code is just... typing some things and having R parrot
them back to you.

```{r}

1

```

```{r}

"Hello"

```

This also works for mathematical equations and other operations, but R
will complete the operation first and then print out the answer.

```{r}

1 + 4

```

If you put multiple expressions in the same code chunk, you'll see the
results underneath, one after the other.

```{r}

"Hello"

3 + 4

```

## Creating Variables

Eventually, though, you will want to store some of these things in the
computers memory, or the R "environment." When you store a value or the
result of an operation to the environment, you're creating a variable.

In R, to create a variable, you start by typing the name of the
variable, and then you use an assignment arrow (`<-`, or a `<` followed
by a `-`) to assign some data to that variable.

```{r}

a <- 1

b <- "one"

```

Note: creating a variable is kind of invisible. The new variable should
show up in your Environment tab in RStudio, but R doesn't print anything
out under the code chunk. If you also want to just see what got stored
in the variable, you can type the variable name on its own.

```{r}

a

b

```

There are basic rules about what you can use as a variable name in R.
Generally, you can't use a '-' or a space in the variable name, you
can't start the name with a number, and you can't include spaces. It is
common in R code to replace spaces with periods (.) or underscores (\_).

These rules, however, can be circumvented by surrounding the name with
back tick marks (\`\`). While sometimes this can be useful, it also
makes things harder because you will always have to use those ticks to
call the variable.

```{r}

# Anything after a '#" on a line is a comment. It will be ignored by the processor.
# I have commented out the examples that would lead to an error because the error
# prevents any additional code from being processed.

# 1a <- "doesn't work"

# 1-a <- "doesn't work"

# 1 a <- "doesn't work"

`1a` <- "does work"

`1-a` <- "does work"

`1 a` <- "does work"

`1a`

`1-a`

`1 a`

```

## Data Types

Anything you store as a variable will be given a data type. The core
data types for simple data objects are:

-   character
-   numeric
-   logical (true/false)

There are some more specific data types that build on these, like
integer and date, but these simple ones, which are technically classes,
are the most common. You can use `class()` to check the simple type of a
variable. The `typeof()` function can show the more detailed data types.

```{r}

class(a)

class(b)

typeof(a)

typeof(b)

```

> ### Sidebar: Using Functions
>
> The way functions work in R is pretty standard for programming
> languages. The name of the function is followed immediately by the
> left parenthesis `(`. Any inputs to the function come next, and then
> you need to close the function with the right parenthesis `)`. You can
> (and should!) write your own functions, but we can cover that later.

## Vectors

In R, the main way of collecting data items into a group is to create a
vector. Vectors are specifically for items that are all the same data
type, though.

To create a vector, use the syntax `c()`. This is the "combine"
function. Inside the parentheses, just list items separated by commas.

```{r}

sample_vector <- c(1, 3, 5, 7)

sample_vector

class(sample_vector)

typeof(sample_vector)

```

Note: the class of a vector is the class of the items in the vector.
Same with type.

Inside the combine function, you can mix things you type with variables
you've already created.

```{r}

sample_vector_2 <- c(b, "three", "five", "seven")

sample_vector_2

class(sample_vector_2)

typeof(sample_vector_2)

```

But you can't have a vector that mixes data types. What do you think
will happen with the following vector?

```{r}

sample_vector_3 <- c(1, "three", TRUE)

```

We didn't get an error, but we were trying to make a vector from a
number, a string, and a boolean. Let's look at what got stored in the
vector.

```{r}

sample_vector_3

class(sample_vector_3)

typeof(sample_vector_3)

```

R did what a lot of programming languages will do in this situation. It
converted the data types on the fly. Everything can be represented by a
string, so R used the character data type for everything.

## Lists

What if you do have data elements of different types and you want to
combine them for some reason? R uses lists to combine mixed type
objects.

The great thing about vectors is that, when you know everything in a
vector is the same type, you know that you can perform the same kind of
operation on every item. That's not true with lists, but sometimes
they're still useful.

You can create a list with the `list()` function.

```{r}

sample_list <- list(1, "three", TRUE)

sample_list

class(sample_list)

typeof(sample_list)

```

Each item in the list has an identifier, signified by two pairs of square brackets
`[[]]`. They are just numbered in order.

## Data Frames

Data frames are the main data structure you'll be using for any work
that uses tabular data. In essense, a data frame has rows and columns.
The columns are all vectors, which means each column has to have
elements of the same data type. But data frames also let you access data
by rows.

Many times, you'll just load data into a data frame from a data file,
like a spreadsheet. But you can also create data frames manually. An
easy way is to build vectors and then give them names as part of the
`data.frame()` function.

```{r}

sample_dataframe <- data.frame(columnA = sample_vector, 
                               columnB = sample_vector_2, 
                               columnC = c("a", "b", "c", "d"))

sample_dataframe


```

What happens if you try to create a data frame with vectors of different
sizes?

```{r}

sample_dataframe_2 <- data.frame(columnA = sample_vector, 
                               columnB = sample_vector_2, 
                               columnC = c("a", "b", "c"))

sample_dataframe_2

```

It makes sense that you'd get an error, trying to use two vectors of
length four and one vector of length three. But be careful! Look at the
next example.

```{r}

sample_dataframe_3 <- data.frame(columnA = sample_vector, 
                               columnB = sample_vector_2, 
                               columnC = c("a", "b"))

sample_dataframe_3

```

No error! What happened?

Because the final column has a length of two, and the others are length
four, R decided to use two copies of the vector in column C. If that's
not what you want, you should be very careful to specify every value in
every column.

Another thing to note about data frames is that they are more particular
about column names than our variable samples above. Even if we use the
\` character to surround column names, we still can't have column names
with spaces or dashes, or have ones that start with a number.

```{r}

sample_dataframe_4 <- data.frame(`column A` = sample_vector, 
                               `column-B` = sample_vector_2, 
                               `3column` = c("a", "b", "c", "d"))

sample_dataframe_4

```

Data frames are the basic data structure used for tabular data. Very
soon, we're going to be working with a more sophisticated version of a
data frame called a `tibble`. This is the type of data frame that is
created when using the `tidyverse` packages to do data science.

Sometimes, for small data sets, you may just want to be able to type out
a comma-separated table and load it in as a data frame. The tidyverse
way of doing that is `tribble()`. A couple of notes: \* the column names
in the first row need to begin with a `~`; that won't be included as
part of the name, though. \* each line also needs to end with a comma,
except for the last one

```{r}

sample_tibble <- tribble(
  ~`column A`, ~`column-B`, ~`3column`,
  1, "one", "a",
  3, "three", "b",
  5, "five", "c",
  7, "seven", "d"
)

sample_tibble

```

Bonus - the tibble that is created uses the actual names we specified!
But even if you're just starting with the basic data frame, all of the
tidyverse operations will still work fine.

## Bonus: Named Vectors

Somewhere in between a vector and a data frame is a named vector. It's a vector
where each item also has a name. It's hard to explain generally why you might need
something like this, but they can come in handy in certain parts of data work.

Creating a named vector is like creating a data frame, but you just assign one
value to each name.

```{r}

sample_named_vector <- c("alpha" = 1, "beta" = 2, "gamma" = 3, "delta" = 4)

sample_named_vector

class(sample_named_vector)

typeof(sample_named_vector)

```

See how even with strings being used to name the elements, the vector is still a 
numeric vector at its core.

## Bonus Bonus: Named Lists

A named list is like something that might be called an object in another programming
language. You can store whatever you want in each name, and they will all be contained
within the named list.

```{r}

sample_named_list <- list("alpha" = 1, "beta" = "three", "gamma" = TRUE)

sample_named_list

class(sample_named_list)

typeof(sample_named_list)

```

When R prints the named list, each name shows up after a dollar sign `$` (e.g., `$alpha`), 
instead of having a position-based identifier (e.g., `[[1]]`).

## Accessing Data: Vectors

Okay, once we have combined data elements into larger structures like vectors,
we need to know how to get them back out again. With vectors, you can use square
brackets `[]` and the position of the value. In R, unlike Python, you count the
position starting with 1.


```{r}

sample_vector[1]

sample_vector_2[2]

```

Using a negative position excludes the item at that position.

```{r}

sample_vector[-4]

sample_vector_2[-2]

```


If you want multiple items, you can specific a vector of positions.

```{r}

sample_vector[c(1,4)]

sample_vector[c(-1,-4)]

```

What if you want all of the numbers up to, or after, a certain position? R uses the
colon `:` character to specify a range of integers. This also works if you want to 
exclude a range.

Order matters for positive number ranges.

```{r}

sample_vector[1:3]

sample_vector[3:1]

sample_vector[-2:-3]

sample_vector[-3:-2]

```

This way of including and excluding items is basically the same as saying yes or
no to different items in the vector. We can actually use a vector of boolean 
values to do the same thing. (This may not seem useful right now, but bear with me.)

```{r}

sample_vector[c(TRUE, FALSE, TRUE, FALSE)]

# Alternate way of typing TRUE and FALSE

sample_vector_2[c(F, F, T, T)]

```

Now lets say we want to subset a vector by some kind of test or condition - all
of the numbers above a certain value, for examples.

First, let's test the vector elements to see if they meet a condition, like
being greater than 4.

```{r}

sample_vector > 4

```

So, what is happening here? We typed an operation, and R evaluated that operation
and printed the result. We asked if a vector was greater than 4, but what R
calculated was whether each item in the vector is greater than 4. This is our
first example of a **vectorized operation** - taking an operation that is meant to work
on a single item and applying it to a whole vector instead. This is a very efficient
way of performing data operations, and it's very common in R to be able to do this.

In vectorized operations like this, you start with a vector, perform some operation
or test on each item, then return a vector of the results for each item. What we
have above is a vector saying, "1 is not greater than 4, 3 is not greater than 4,
5 is greater than 4, 7 is greater than 4."

But we know we can use a vector of boolean values to subset a list. So, let's take
our `> 4` test and insert it into our square brackets to subset the original vector.

```{r}

sample_vector[sample_vector > 4]

```

This is saying "Subset sample_vector based on the results of testing each element
in the vector on whether it is greater than 4."

This is a bit redundant, and you have to remember this confusing pattern to be able 
to use it. It does allow for some funny (useful?) things, though, like mixing vectors.

```{r}

sample_vector_2[sample_vector > 4]

```


## Accessing Data: Named Vectors

If you have a named vector, you can still use position and logic to subset the vector,
but you can also use the names. This is a great feature of named vectors, since
it is much harder to get a name wrong than a numerical position.

```{r}

sample_named_vector[c(1,3)]

sample_named_vector[c("beta","delta")]

sample_named_vector[sample_named_vector == 3]

```

## Accessing Data: Lists

For unnamed lists, we're back to using position, or maybe equality. But you 
do have two options. You can take the list and limit it to certain items, or
you can remove an item from the list and get it back in its original form

```{r}

sample_list[2]

class(sample_list[2])

sample_list[[3]]

class(sample_list[[3]])

```

In the first example, `sample_list[2]`, we are saying, "Include the 2nd element only,
but still give me a list." This actually returns a list of one element, and that element
has gotten a new automatic identifier with a position of 1 (`[[1]]`).

In the second example, we use two sets of square brackets, which tells R we want
to drill all the way down to the original element. It prints out `TRUE`, and we 
can confirm that the returned value has a class of "logical," or boolean.

We can still ask for multiple items using the first method, but not the second.
We'd have to get the individual items and then reconsititute the list.

```{r}

sample_list[c(1,3)]

# doesn't work
# sample_list[[c(1,3)]]

list(sample_list[[1]], sample_list[[3]])

```

Now, our previous trick of testing vector elements based on whether they're greater
than some number is harder with a list. Each list element could be a different class.

One thing we could try is to limit the list just to elements of a certain class. There
are a bunch of functions in R that test if an element is a particular data type, like
`is.character()` and `is.numeric()`. Let's try one on a list.

```{r}


is.numeric(sample_list)
is.character(sample_list)
is.logical(sample_list)


```

Unfortunately, that approach doesn't work. Functions may be vectorized, meaning that
they work over elements in a vector, but lists by their nature are so complex that
functions will normally treat them as a whole object. We can iterate over list
items, but that's a more advanced topic than we'll cover here.


## Accessing Data: Named Lists

Square brackets will also work here, and the rules are the same as with positions.
You just have to surround the item name in quotes.

```{r}

# prints both the identifier and the value; identifier has not changed because
# identifier isn't tied to position anymore
sample_named_list["alpha"]

# prints just the original item, the string "three"
sample_named_list[["beta"]]

# prints a named list, reduced to just two named items - alpha and gamma
sample_named_list[c("alpha","gamma")]

# doesn't work
# sample_named_list[[c("alpha","gamma")]]

```

But just like we saw with unnamed lists, R is actually giving us useful information when it 
prints out the named list. It uses `$` before the name of each list item, and that
is also how we can access those items directly. This syntax returns the item itself,
like the double square brackets.

```{r}

sample_named_list$alpha

sample_named_list$beta

```

If you have weird names, the dollar sign can be tricky. But you can still use
tick marks to surround the weird name. 

```{r}

sample_named_list_2 <- list(`weird 1` = 5, `2weird` = 6)

sample_named_list_2$`weird 1`

sample_named_list_2[["weird 1"]]

```

## Accessing Data: Data Frames

Now that you're starting to learn about how R represents different kinds of
data structures, let's look at our environment. We have two sections: "Data" at
the top and "Values" at the bottom. In the Data section, we have data frames
and lists. In the Values section we have vectors and individual data elements.

Focusing in on the Data section, use the arrow on the left to expand a data frame.
Then expand a list. In each case we see the different elements of the structure,
one per row, sometimes with a name and sometimes without. What else do you notice?

Under both lists and data frames, we see a `$` at the beginning of every row.

We've already seen we can access the named elements of a list using `$`. It turns
out, the same is true for data frames.

```{r}

sample_dataframe$columnA

```


Can you guess why that is? What does that say about data frames?

...

..

.

A data frame is a list! A named list!

What makes it different? What are the constraints for a data frame that don't exist
for a list?

...

..

.

A data frame is a (named) list of vectors! Each column is a separate element
in the list, and each column must be a vector. But different columns can have
different data types, just like different elements in a normal list.

Because a data frame is a list, you can use either position or name to access columns.
And you can use either `$` or `[]`.


```{r}

sample_dataframe[1]
sample_dataframe["columnB"]
sample_dataframe[c(1,3)]
sample_dataframe[-2]

```

Just like with a list, using the dollar sign with a data frame returns a vector. So does `[[]]`, using either position or name.

```{r}

sample_dataframe$columnB
sample_dataframe[[1]]
sample_dataframe[["columnC"]]

```


Just like with lists, you can't pass a vector of two numbers to `[[]]`.

```{r}

sample_dataframe[c(1,3)]

```



Data frames also have an option I think of as "matrix" syntax. In square brackets `[]`, you can actually use two numbers separated by a comma to select a specific row and column, respectively.

```{r}

sample_dataframe[1,2]

```

If you want the entire column, you can skip the number for the row, and vice versa.

```{r}

sample_dataframe[,2]

sample_dataframe[3,]


```

In the first case, grabbing the column, you get just the vector and not a data frame. But grabbing the row, you get a data frame. Why might that be?

...

..

.

Columns are strictly typed, because they are vectors by definition. So you always 
know the elements in a column are a vector at the core, and that's what this syntax
returns. But row elements could all be different types, so they can't come back as
a vector. The safe thing to do is to keep them in a data frame. You could imagine
getting a list back, but a data frame is actually better because we know that each
column has the same number of rows, and we might want to do more data frame operations on those rows, especially if we requested a range of rows.

```{r}

sample_dataframe[c(1,3),]

```

We could keep going with this syntax and how you can subset more creatively using
logical tests, but I think the tidyverse has better methods than this. 

Also, it is possible to set row names for data frames. You could identify a column
that acts like an index of the data, one that has a unique identifier for each row.
That column could be used for row names, and then you could use name-based subsetting
like we do for columns. In practice, though, you don't often want to isolate rows
by their identifier. More commonly you want to isolate them based on patterns in the
data. We're going to look at the `tidyverse` solutions now, and you can still do both:
identify a particular row based on its identifier, and also look for groups of rows
that meet some kind of data condition.

Let's move on.
